---
title: "Managing Data"
output: github_document
---


```{r setup, include = FALSE, echo = FALSE}
library(tidyverse)
options(digits = 3)
```

```{r load_customer_data}
customer_data <- read_rds('data/customer_data.rds')
skimr::skim(customer_data)
```


Treating the age and income variables

- 001: Included in rent or condo fee
- 002: Included in electricitypayment
- 003: no charge or gas not used

```{r dealing_with_sentinel_values}
customer_data <- customer_data %>% 
    mutate(
        age = na_if(age, 0),
        # convert negative incomes to NA
        income = ifelse(income < 0, NA, income),
        # create 3 indicator variables of gas usage
        gas_with_rent = (gas_usage == 1),
        gas_with_electricity = (gas_usage == 2),
        no_gas_bill = (gas_usage == 3),
        gas_usage = ifelse(gas_usage < 4, NA, gas_usage)
    )
```


Missing values with `vtreat`

```{r missing_value_functions}
na_counts <- function(df) {
    map_int(df, ~sum(is.na(.x)))
}

na_summary <- function(df) {
    map(df, ~sum(is.na(.x)))
}
```

```{r missing_value_counts}
na_counts(customer_data)
```


```{r create_treatment_plan_for_missing_values}
treatment_plan <- vtreat::design_missingness_treatment(
    customer_data,
    varlist = setdiff(colnames(customer_data), c('custid', 'health_ins'))
)
customer_prepared <- vtreat::prepare(
    treatment_plan,
    customer_data
)
customer_prepared
```

```{r summarise_differences_in_nas}
# convert two lists to two rows in a data frame
list(
    original = na_summary(customer_data),
    prepared = na_summary(customer_prepared)
) %>% bind_rows(.id = 'custid')
```



Examining the data treatment

```{r examining_treated_data}
# rows with no housing type defined
ht_missing <- which(is.na(customer_data$housing_type))
customer_data %>% 
    slice(ht_missing) %>% 
    select(custid, is_employed, num_vehicles, housing_type, health_ins) %>% 
    head()

customer_prepared %>% 
    slice(ht_missing) %>% 
    select(
        custid,
        is_employed,
        is_employed_isBAD,
        num_vehicles,
        num_vehicles_isBAD,
        housing_type,
        health_ins
    ) %>% 
    head()
rm(ht_missing)

```

Verify the expected number of vehicles and employment rate

```{r verify_imputed_means}
customer_data %>% 
    summarise_at(vars(num_vehicles, is_employed), mean, na.rm = TRUE)
```

## Data Transformations

### Normalizion

Normalization (or rescaling) is useful when absolute quantities are less meaningful than relative ones.

Example Suppose you are considering the use of income as an input to your insurance model.

The cost of living will vary from state to state, so what would be a high salary in one region could be barely enough to scrape by in another.

Because of this, it might be more meaningful to normalize a customer’s income by the typical income in the area where they live.

This is an example of a relatively simple (and common) transformation.


```{r load_median_income}
# load the lookup table
median_income <- read_rds('data/median_income.rds')
median_income %>% arrange(desc(median_income)) %>%  head(n = 10)

```

Normalize the income

```{r normalizing_the_income}
customer_prepared <- customer_prepared %>% 
    left_join(median_income, by = 'state_of_res') %>% 
    mutate(income_normalized = income / median_income)


customer_prepared %>% 
    select(income, median_income, income_normalized) %>% 
    head()

summary(customer_prepared$income_normalized)
```

Looking at the results you see that customers with an income higher than the median income of their state have an `income_normalized` value larger than 1, and customers with an income lower than the median income of their state have an `income_normalized` value less than 1.

Because customers in different states get a different normalization, we call this a conditional transform.

A long way to say this is that _the normalization is conditioned on the customer’s state of residence_.

We would call scaling all the customers by the same value an unconditioned transform.

Another example, you might be less interested in a customer’s absolute age than you are in how old or young they are relative to a _typical_ customer. Let’s take the mean age of your customers to be the typical age.

```{r}
summary(customer_prepared$age / mean(customer_prepared$age))
```


### Centering and scaling

```{r scaled_age}
mean_age <- mean(customer_prepared$age)
sd_age <- sd(customer_prepared$age)
customer_prepared <- customer_prepared %>% 
    mutate(scaled_age = (age - mean_age) / sd_age)

# typical
customer_prepared %>% 
    select(age, scaled_age) %>% 
    filter(abs(age - mean_age) < sd_age) %>% 
    head()

# non-typical
customer_prepared %>% 
    select(age, scaled_age) %>% 
    filter(abs(age - mean_age) > sd_age) %>% 
    head()
```

```{r scale_age_cleanup}
rm(mean_age, sd_age)
```


Use the built in `scale` function

```{r compare_with_base_scale}
head(scale(customer_prepared$age))
head(customer_prepared$scaled_age)
    
```

```{r scaled_features}
data_scaled <- customer_prepared %>% 
    select(age, income, num_vehicles, gas_usage) %>% 
    scale(center = TRUE, scale = TRUE)
summary(data_scaled)
# means
(scaled_means <- attr(data_scaled, 'scaled:center'))
# sds
(scaled_sds <- attr(data_scaled, 'scaled:scale'))
```


Treating new data

```{r scaling_using_vtreat}
new_data <- customer_data %>% 
    sample_n(100)
# clean using the treatment plan from the original dataset
new_data <- vtreat::prepare(treatment_plan, new_data)
# scale using the centers and scaling factors from the original dataset
new_data_scaled <- new_data %>% 
    select(age, income, num_vehicles, gas_usage) %>% 
    scale(center = scaled_means, scale = scaled_sds)

summary(new_data_scaled)
```

### Log transformations for skewed and wide distributions

```{r plot_log_income, fig.width = 10, fig.height = 7}
customer_prepared %>% 
    ggplot(aes(x = income)) +
    geom_density() +
    scale_x_log10(labels = scales::dollar)

summary(customer_prepared$income)
```


Signed log

```{r}
signed_log10 <- function(x) {
    # -1 to 1 is mapped to 0
    ifelse(abs(x) <= 1, 0, sign(x) * log10(abs(x)))
}
```


### Test and training splits

