---
title: "Data engineering and data shaping"
output: github_document
---


```{r, setup}
library(tidyverse)
options(digits = 3)
```

### Choosing a subset of columns

```{r iris_summary}
skimr::skim(iris)
```


```{r plot_iris_petal_measurements}
ggplot(iris, aes(
    x = Petal.Length,
    y = Petal.Width,
    shape = Species,
    colour = Species
)) +
geom_point(size = 2) +
ggtitle('Petal dimensions by iris species: all measurements')
```

Suppose we are assigned to generate a report on only petal length and petal width, by iris species, for irises where the petal length is greater than 2.

To accomplish this, we need to select a subset of columns (variables) or a subset of rows (instances) from a data frame.

#### Base R

```{r iris_with_base_r}
columns_we_want <- c("Petal.Length", "Petal.Width", "Species")
rows_we_want <- which(iris$Petal.Length > 2)
# before
head(iris)
# include drop=False to ensure that a data frame is always returned
iris_base <- iris[rows_we_want, columns_we_want, drop = FALSE]
# after
iris_base

```

#### Data.Table

```{r iris_with_data_table}
library(data.table)
iris_data <- as.data.table(iris)
columns_we_want <- c("Petal.Length", "Petal.Width", "Species")
rows_we_want <- iris$Petal.Length > 2
iris_data <- iris_data[rows_we_want, ..columns_we_want]
head(iris_data)

```


#### dplyr

```{r iris_with_dplyr}
iris_data <- iris %>% 
    select(starts_with('Petal'), Species) %>% 
    filter(Petal.Length > 2)
head(iris_data)
```

### Removing records with incomplete data

```{r msleep}
head(msleep)
? msleep
```

#### Base R

```{r}
clean_base <- msleep[complete.cases(msleep),, drop = FALSE]
summary(clean_base)
# or
clean_base <- na.omit(msleep)

```


### Ordering Rows


```{r}
purchases <- tibble(
    day = c(1,2,2,1,2,1),
    hour = c(9, 9, 11, 13, 13, 14),
    n_purchase = c(5, 3, 5, 1, 3, 1)
)
```

#### Base R

```{r}
order_index <- with(purchases, order(day, hour))
  
purchases_ordered <- purchases[order_index, , drop = FALSE]
purchases_ordered$running_total <- cumsum(purchases_ordered$n_purchase)

purchases_ordered
```


#### Data Table

```{r}
library("data.table")

DT_purchases <- as.data.table(purchases)

order_cols <- c("day", "hour")
setorderv(DT_purchases, order_cols)

DT_purchases[, running_total := cumsum(n_purchase)]
```


#### dplyr

```{r}
purchases %>%
    arrange(day, hour) %>%
    mutate(running_total = cumsum(n_purchase))
```


### Advanced Ordering

Cumulative sum of sales to be per-day

#### Base R

```{r}
# sort
order_index <- with(purchases, order(day, hour)) 
purchases_ordered <- purchases[order_index, , drop = FALSE]
# split into a list of groups
data_list <- split(purchases_ordered, purchases_ordered$day)
# apply cumsum to each group
data_list <- lapply(
    data_list,
    function(di) {
        di$running_total <- cumsum(di$n_purchase)
        di
    }
)
# put the results back together in a single data frame
purchases_ordered <- do.call(base::rbind, data_list)
# remove any annotations
rownames(purchases_ordered) <- NULL

purchases_ordered
```

You don't want to know how `data.table` does it

#### dplyr

```{r}
purchases %>%
    arrange(day, hour) %>%
    group_by(day) %>%
    mutate(running_total = cumsum(n_purchase)) %>%
    ungroup()
```


### Transformations

```{r convert_date_components_to_strings}
library(lubridate)
datestr = function(day, month, year) {
    paste(day, month, year, sep="-")
}


```

#### Data Table

```{r}
DT_airquality <- as.data.table(airquality)

DT_airquality[, date := dmy(datestr(Day, Month, 1973))]
head(DT_airquality[, c('Ozone', 'date')])

```



#### dplyr

```{r}
DP_airquality <- airquality %>% 
    mutate(date = dmy(datestr(Day, Month, 1973))) %>% 
    select(Ozone, date)
head(DP_airquality)

```


Fix the missing values

```{r air_quality_missing_data_table}
DT_airquality[, OzoneCorrected := zoo::na.locf(Ozone, na.rm = FALSE)]
summary(DT_airquality)
```

```{r interpolate_last_observation_carried_forward}
zoo::na.locf(DP_airquality$Ozone, na.rm=FALSE)
```



```{r plot_air_quality_carried_forward}
DP_airquality <- DP_airquality %>% 
    mutate(OzoneCorrected = zoo::na.locf(Ozone, na.rm = FALSE))
DP_airquality %>% 
    ggplot(aes(x = date, y = Ozone)) +
    geom_point() +
    geom_line(aes(y = OzoneCorrected)) +
    ggtitle(
        'New York ozone readings, May 1 to Sept 30 1973',
        subtitle = '(corrected)'
    ) +
    labs(x = NULL)

```


### Aggregation

#### R Base

```{r aggregate_iris_base}
iris_summary <- aggregate(
    cbind(Petal.Length, Petal.Width) ~ Species,
    data = iris,
    FUN = mean
)
print(iris_summary)
```

#### Data Table

```{r aggregate_iris_data_table}
iris_data.table <- as.data.table(iris)
iris_data.table <- iris_data.table[, .(
    Petal.Length = mean(Petal.Length),
    Petal.Width = mean(Petal.Width)
), by = .(Species)]
```


#### dplyr

```{r aggregate_iris_tidy}

iris_summary <- iris %>% group_by(Species) %>% 
    summarize_at(
        vars(Petal.Length, Petal.Width),
        mean,
        .groups = 'drop_last'
    )


ggplot(mapping=aes(x = Petal.Length, y = Petal.Width, shape = Species, colour = Species)) +
    geom_point(data = iris, alpha = 0.5) +
    geom_point(data = iris_summary, size = 5) +
    ggtitle("Average Petal dimensions by iris species\n(with raw data for reference)")

```




