---
title: "K-Nearest Neighbors"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE
)
options(digits = 3)
library(tidyverse)

accuracy <- function(cm) {
    sum(diag(cm)) / sum(cm)
}
```

## Stock Market Data

```{r load_stock_market}
stock_market <- read_rds('data/stock_market.rds')
```

```{r stock_market_counts_for_year}
stock_market %>% 
    count(Year)
```

```{r stock_market_counts_for_direction}
stock_market %>% 
    count(Direction)
```



```{r stock_market_train_test}
# everrything before 2005
train <- stock_market %>% 
    filter(Year < 2005)
# 2005
test <- stock_market %>% 
    filter(Year >= 2005)
```


```{r stock_market_lag1_lag2}
stock_market %>% 
    select(starts_with('Lag')) %>% 
    cor() %>% 
    round(digits = 2)
```


### Stock Market KNN

1. A matrix containing the predictors associated with the training data,  labeled train.X below.
2. A matrix containing the predictors associated with the data for which we wish to make predictions, labeled test.X below.
3. A vector containing the class labels for the training observations, labeled train.Direction below.
4. A value for K, the number of nearest neighbors to be used by the classifier.



```{r stock_market_knn_data}
train_x <- train %>% 
    select(Lag1, Lag2) %>% 
    data.matrix()
test_x <- test %>% 
    select(Lag1, Lag2) %>% 
    data.matrix()
train_y <- train %>% 
    pull(Direction)

test_y <- test %>% 
    pull(Direction)

```

```{r stock_market_knn_1}
# k = 1
knn_1 <- class::knn(
    train_x,
    test_x,
    train_y,
    k = 1
)
# how well does <2005 predict 2005
cm_1 <- table(`<2005` = knn_1, `2005` = test_y)
cm_1
accuracy(cm_1)
```

Only 50% of the observations were correctly predicted. `k=1` is far too flexible


```{r stock_market_knn_3}
knn_3 <- class::knn(
    train_x,
    test_x,
    train_y,
    k = 3
)
cm_3 <- table(`<2005` = knn_3, `2005` = test_y)
cm_3
accuracy(cm_3)

```

## Caravan Insurance


```{r load_caravan_data}
caravan <- read_rds('data/caravan.rds')
prop.table(xtabs(~Purchase, data = caravan))
```

```{r caravan_missing}
caravan %>% 
    select(-Purchase) %>% 
    summarise_all(
        ~sum(is.na(.))
    ) %>% 
    pivot_longer(everything(), names_to = 'Var', values_to = 'NAs') %>% 
    filter(NAs > 0) %>% 
    arrange(desc(NAs))
```


```{r caravan_train_test}
caravan_m <- scale(select(Caravan, -Purchase))
purchase <- pull(caravan, Purchase)
# rgroup <- sample(1:nrow(caravan), 1000, replace = F)
rgroup <- 1:1000
apply(caravan_m, 2, var)
round(apply(caravan_m, 2, mean), 2)
train_x <- caravan_m[-rgroup,]
train_y <- purchase[-rgroup]
test_x <- caravan_m[rgroup,]
test_y <- purchase[rgroup]
rm(caravan_m, rgroup)
```


```{r caravan_knn_1}
table(test_y)
set.seed(1)
knn_1 <- class::knn(
    train_x,
    test_x,
    train_y,
    k = 1
)
cm_1 <- table(actual = test_y, predicted = knn_1)
cm_1
# accuracy is not good for rare data
accuracy(cm_1)
# error rate
prop.table(table(test_y != knn_1)) * 100
# you could get this down to 6% by always predicting No
prop.table(table(test_y)) * 100
```


```{r knn_1_row_proportions}
cm_1
table(test_y)
prop.table(cm_1, margin = 2)
```

About double the rate predicted by chance


```{r caravan_knn_3}
knn_3 <- class::knn(
    train_x,
    test_x,
    train_y,
    k = 3
)
cm_3 <- table(actual = test_y, predicted = knn_3)
prop.table(cm_3, margin = 2)
```

```{r caravan_knn_n}
map(1:5, ~prop.table(
    table(actual = test_y, predicted = class::knn(train_x, test_x, train_y, k = .)),
    margin = 2
))
```

It appears that KNN is finding some real patterns in a difficult data set!
