---
title: "Time Series Smoothing"
output: html_document
editor_options: 
    chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    dpi = 100,
    fig.width = 11,
    fig.height = 6,
    message = F,
    error = F,
    warning = F,
    cache = F
)
options(digits=4)
library(tidyverse)
library(lubridate)

date_rotation <- ggExtra::rotateTextX(angle = 45, vjust = 1, hjust = 1)

```

## Exponential Smoothing

Start with an additive model with constant level `m` and no seasonality:

$$
X_t = m + W_t
$$

Where $W_t$ is the irregular component.

To forcast $X_{n+1}$ given $x_n, x_{n-1},x_{n-2}\ldots$ we use **exponential smoothing**:

$$
\hat x_{n+1} = c_0 x_n + c_1 x_{n-1} + c_2 x_{n-2} + c_3 x_{n-3} \ldots
$$

Where the weights are

$$
c_i = \alpha(1-\alpha)^i,i = 0,1,2,\ldots
$$

Where &alpha; is a parameter to be estimated and the weights sum to 1. so:

$$
\hat x_{n+1} = \alpha x_n + \alpha(1-\alpha) x_{n-2} + \alpha(1-\alpha)^2 x_{n-2} + \alpha(1-\alpha)^3 x_{n-3}\ldots
$$

### Calculating weights

Calculate the wieghts $c_i = \alpha(1-\alpha)^i$ for $i = 0, 1, 2, 3, 4$ when &alpha; is 0.5 and 0.8

```{r calculating_weights}
calc_weights <- function(a, n) {
    map_dbl(0:n, function(i) a*(1 - a)^i)
}

tibble(
    c = 0:4,
    `0.5` = calc_weights(0.5, 4),
    `0.8` = calc_weights(0.8, 4)
)
```

The weight given to the current observation is $c_0$, which is equal to &alpha;.

Thus more weight is given to the current observation when &alpha; = 0.8 than when &alpha; = 0.5.


### Simulating exponential smoothing

Smoothing can be rewritten like this:

$$
\hat x_{n+1} = \alpha x_n + (1-\alpha) \hat x_n
$$

Hence the 1-step ahead forecast of $X_{n+1}$ is a combination of $x_n$, (the observed value of $X_n$ and $\hat x_n$ (the 1-step ahead forecast of $X_n$)

```{r simulating_exponential_forecasting}
X <- c(19.4, 18.9, 18.9, 17.8)
n <- length(X)
Y <- numeric(n)
alpha = 0.6
# seed
Y[1] <- X[1]
for(i in 2:n) {
    # observed value + one-step ahead forecast
    Y[i] <- alpha * X[i-1] + (1-alpha) * Y[i-1]
}
tibble(
    index = 1:n,
    X = X,
    Y = Y
)
```

### Predict tomorrow's temperature

```{r predicting_tomorrows_temperature}
# later temperatures
X <- c(18.9, 17.8, 19.4, NA)
n <- length(X)
Y <- numeric(n)
alpha <- 0.6
# the 1-step ahead is currently set to
Y[1] <- 20.26312
for (i in 2:n) {
    Y[i] <- alpha * X[i-1] + (1 - alpha) * Y[i-1]
}

# 29:(29+n)

tibble(
    index = 29:(29+n-1),
    X = X,
    Y = Y
)
```

## Example: Chemical Process Data

Load the data


```{r load_chemical_process_data}
chemical = read_csv(
    'data/chemical_process.csv',
    col_types = cols(
        index = col_integer(),
        temperature = col_integer()
    )
)
```



Fit smoothing model

```{r chemical_smoothing_model}
# find the value of alpha that minimizes the sse
chemical_fit <- forecast::ses(
    chemical$temperature,
    # want to optimize intial values as well as the alpha smoothing parameter
    initial="optimal"
)
coef(chemical_fit$model)
```

Assessing the accuracy of the forecasts

```{r chemical_smoothing_fitted_and_residuals}
# add fitted and residuals
chemical <- chemical %>% 
    add_column(forecast = fitted(chemical_fit)) %>% 
    mutate(resid = temperature - forecast)
```



```{r chemical_smoothing_accuracy}

chemical %>% 
    summarise(
        SSE = sum(resid^2)
    ) %>% 
    # n-k = 100-1 since we are only estimating alpha
    mutate(RMSE = sqrt(SSE/99))
```


The accuracy of the forecast for each measurement can be assessed by the difference between the observed value and the forecast.

This difference is called the forecast error and is denoted $\epsilon_t$ . 

The first value is zero because the initial value was chosen to be the temperature initially observed.

A convenient way of summarizing the overall accuracy of the forecasts by a single number is to add up the squares of the forecast errors

```{r plot_chemical_smoothing}
chemical %>% 
    pivot_longer(temperature:forecast, names_to = 'series', values_to = 'temperature') %>% 
    ggplot(aes(index, temperature, colour = series)) +
    geom_line()
```



## Holts exponential smoothing method

This is an extension of simple exponential smoothing which is approproate for time series with a linear trend

Start with an additive non-seasonal model with a *linear trend* component

$$
X_t = m + b t + W_t
$$

where b is the slope of the trend component $m_t = m + bt$. Now

$$
X_{t+1} = m + b(t+1) + W_{t+1} \\
= (m + bt) + b + W_{t + 1} \\
= m_t + b + W_{t + 1}
$$

So now the 1-step ahead foreacast for $X_t+1$ is obtained using simple exponential smoothing is an estimate of $m_t$, the level at time t, so that

$$
\hat x_{t + 1} = \hat m_t
$$

Since $W_t+1$ has mean zeros, the expected value of $X_{t+1}$ is $m_{t} + b$. Hence a better forecast is

$$
\hat x_{t+1} = \hat m_t + \hat b
$$

where $\hat b$ is an estimate of the slope. In the example above it would mean adding a bit on in line with recent increase in temperatures.

Holt’s exponential smoothing method provides a way of estimating both the level $m_t$ and the slope `b` at each time point. This is achieved by a smoothing method very similar to that used in simple exponential smoothing.

The main difference between Holt’s method and simple exponential smoothing is that the smoothing is now controlled by two parameters - one parameter &alpha; for the estimate of the level $m_t$ at the current time point, and one parameter &beta; for the estimate of the slope b.

The values of both parameters, &alpha; and &beta;, lie between 0 and 1. As with simple exponential smoothing, values of the parameters close to 0 mean that little weight is placed on the most recent observations.

On the other hand, values close to 1 mean that much weight is placed on the most recent observations.

## Example: Central England Temperatures, 1901-2004


```{r load_cet}
temp_df <- read_csv(
    'data/central_england_temperatures_1659_2004.csv',
    col_types = cols(
        year = col_integer(),
        temperature = col_double()
    )
)
```


```{r cet_etl}
temp_df <- temp_df %>% 
    # we just want 1901-2004
    filter(year > 1900) %>% 
    mutate(
        index = make_date(year)
    ) %>%
    select(index, temperature)
```


```{r cet_summarise}
temp_df %>% 
    summarise(min = min(index), max = max(index))
```


```{r cet_plot, fig.width = 12}
temp_df %>% 
    ggplot(aes(index, temperature)) +
    geom_line() +
    scale_x_date(date_breaks = '5 years', date_labels = '%Y') +
    labs(x = NULL) +
    ggtitle('Annual average Central England temperatures (C), 1901-2004')
```

The data are annual, so the time series is not seasonal, and the width of the irregular fluctuations does not appear to vary with the level. 

Temperatures appear higher towards the end of the series than at the beginning, so it is not unreasonable to assume that there might be an increasing trend.

```{r cet_ses_model}
model_ses <- forecast::ses(temp_df$temperature, initial = 'optimal')
coef(model_ses$model)
```

```{r cet_holt_model}
model_holt <- forecast::holt(temp_df$temperature, initial = 'optimal')
coef(model_holt$model)
```



```{r cet_models_fitted_plus_residuals}
temp_df <- temp_df %>% 
    mutate(
        forecast_ses = fitted(model_ses),
        resid_ses = resid(model_ses),
        forecast_holt = fitted(model_holt),
        resid_holt = resid(model_holt)
    )
```

```{r cet_2004}
temp_df %>% 
    slice(nrow(temp_df))
```


Compute SSE for each model

```{r cet_model_accuracy}
temp_df %>% 
    summarise_at(vars(starts_with('resid')), ~sum(.^2))

```

Plot each model

```{r cet_model_plot, fig.width = 12}
temp_df %>% 
    select(index, temperature, starts_with('forecast')) %>% 
    pivot_longer(temperature:forecast_holt, names_to = 'series', values_to = 'temperature') %>% 
    ggplot(aes(index, temperature, colour = series)) +
    geom_line() +
    scale_x_date(date_breaks = '5 years', date_labels = '%Y') +
    labs(x = NULL)
```


```{r cet_predictions, fig.width = 13}
predict_ses <- as_tibble(predict(model_ses))
predict_holt <- as_tibble(predict(model_holt))
n <- nrow(predict_ses)

predictions <- tibble(
    index = seq(make_date(2005, 1, 1), by = 'year', length.out = n),
    temperature = rep(NA, n),
    forecast_ses = predict_ses$`Point Forecast`,
    forecast_holt = predict_holt$`Point Forecast`
)
```


```{r cet_plot_predictions}
temp_df %>% 
    select(index, temperature, starts_with('forecast')) %>% 
    bind_rows(predictions) %>% 
    pivot_longer(temperature:forecast_holt, names_to = 'series', values_to = 'temperature') %>% 
    ggplot(aes(index, temperature, colour = series)) +
    geom_line() +
    scale_x_date(date_breaks = '5 years', date_labels = '%Y') +
    scale_y_continuous(limits = c(8, 11)) +
    labs(x = NULL)
```

## Holt-Winters exponential smoothing

Holt’s exponential smoothing method applies only to non-seasonal time series.

A further extension of exponential smoothing, that includes both a linear trend component and seasonality, is known as Holt–Winters exponential smoothing.

This method is appropriate for time series that can be described by an additive model with a linear trend component $m + bt$ and a seasonal component $s_t$, that is:


$$
X_t = m + bt + s_t + W_t
$$


So now we have a third smoothing parameter &gamma;

This parameter adjusts the estimate of the seasonal component. Optimal values of the three parameters &alpha;, &beta;
and &gamma; may be chosen by minimizing the SSE


## Forecasting monthly temperatures


```{r load_cet_monthly}
temp_df <- read_csv(
    'data/monthly_temperature_changes_1970_to_2004.csv',
    col_types = cols(
        index = col_date(format='%Y-%m-%d'),
        temperature = col_double()
    )
)
```


```{r cet_year_breaks}
year_breaks <- seq(make_date(1970, 1, 1), by = 'year', length.out = 36)
range(year_breaks)
range(temp_df$index)
```



```{r cet_monthly_plot, fig.width = 11}
temp_df %>% 
    ggplot(aes(index, temperature)) +
    geom_line() +
    scale_x_date(
        breaks = year_breaks,
        date_labels = '%Y',
        expand = c(0.02, 0)
    ) +
    scale_y_continuous(limits = c(-5, 20)) +
    labs(x = NULL) +
    ggtitle('Monthly average temperatures in Central England from January 1970 to December 2004') +
    date_rotation

```

```{r cet_monthly_ts}
cet_ts <- ts(temp_df$temperature, start = c(1970, 1), frequency = 12)
```



```{r cet_monthly_seasonal_plot, fig.width = 12}
forecast::ggseasonplot(
    cet_ts,
    main = 'Seasonal plot of Monthly average temperatures',
    year.labels = FALSE
)
```


```{r cet_monthly_models}
model_ses <- forecast::ses(cet_ts, initial = 'optimal')
model_holt <- forecast::holt(cet_ts, initial='optimal')
model_hw <- forecast::hw(cet_ts, initial = 'optimal')
coef(model_ses$model)
coef(model_holt$model)
coef(model_hw$model)

```

```{r cet_monthly_fitted}
temp_df <- temp_df %>% 
    mutate(
        simple = fitted(model_ses),
        holt = fitted(model_holt),
        holt_winters = fitted(model_hw)
    )
```



```{r plot_cet_monthly_models, fig.width = 10, fig.height = 5}
temp_df %>% 
    filter(year(index) > 2003) %>% 
    rename(observed = temperature) %>% 
    pivot_longer(observed:holt_winters, names_to = 'series', values_to = 'temperature') %>% 
    ggplot(aes(index, temperature)) +
    geom_line(aes(colour = series, linetype = series)) +
    scale_x_date(
        date_breaks = '1 month',
        date_labels = '%b\n%Y'
        
    ) +
    scale_y_continuous(limits = c(0, 20)) +
    labs(x = NULL, y = 'Temperature (C)')
```

