---
title: "GGPLOT2 Workshop Part 1"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ISLR)
```

## Introduction

Here we look at the basic ggplot2 using the faithful dataset. The x axis is how long the eruption lasts and the y axis is the length of the wait.

- Data
- Mapping
- Statistics
- Scales
- Geometries
- Facets
- Coordinates
- Theme

```{r basic_scatterplot}
# basic scatterplot - data is global. colour is an expression not a string
# colour='steelblue'
ggplot(data=faithful, mapping=aes(x=eruptions, y=waiting, colour=eruptions<3)) +
    geom_point()
# data can be given per layer. note that colour is outside of aes
ggplot() +
    geom_point(mapping=aes(x=eruptions, y=waiting), data=faithful, colour='steelblue')

```

There are other geometries

```{r basic_histogram}
ggplot(faithful) +
    geom_histogram(aes(x=eruptions), binwidth=0.2)
```

```{r basic_density_plot}
ggplot(faithful, aes(x=eruptions, y=waiting)) +
    geom_density_2d() +
    geom_point()
```

### Exercises

Modify the code below to make the points larger squares and slightly transparent.
See `?geom_point` for more information on the point layer.

```{r faithful_point_plot}
ggplot(faithful) +
    geom_point(aes(x=eruptions, y=waiting), size=3, shape='square', alpha=0.5)

```

```{r point_shapes_and_sizes}
df2<-data.frame(x=1:5, y=1:25, z=1:25)
s <- ggplot(df2, aes(x = x, y = y))
s + geom_point()

s + geom_point(aes(shape = z), size = 4) + scale_shape_identity()
# While all symbols have a foreground colour, symbols 19-25 also take a
# background colour (fill)
s + geom_point(aes(shape = z), size = 4, colour = "Red") +
    scale_shape_identity()
s + geom_point(aes(shape = z), size = 4, colour = "Red", fill = "Black") +
    scale_shape_identity()

```


```{r faithful_point_plot_2}
ggplot(faithful) +
    geom_point(aes(x=eruptions, y=waiting), alpha=0.5, shape=15, size=2, colour='darkblue')
```

Color the two distributions in the histogram with different colours

```{r faithful_histogram_1}
ggplot(faithful) +
    geom_histogram(
        aes(x=eruptions,fill=eruptions<3),
        colour='black',
        binwidth=0.1
    )
```

```{r faithful_histogram_2}
# try position=identity and alpha=0.2
ggplot(faithful) +
    geom_histogram(
        aes(x=eruptions, fill=waiting<60),
        color='lightgrey',
        position='identity',
        alpha=0.5
    )

```


Separate the two distributions with a line

```{r faithful_two_distributions}
ggplot(faithful) +
    geom_point(aes(x=eruptions, y=waiting)) +
    geom_abline(slope=-40, intercept=190)
```


## Statistics

Statistics represent a transformation of the data prior to plotting.

`geom_bar` uses `stat_count` by default

```{r geom_bar}
ggplot(mpg) +
    geom_bar(aes(x=class))
```

```{r geom_col}
mpg %>% 
    count(class, name = 'count') %>%
    ggplot(aes(x = class, y = count)) +
    geom_col()
```


```{r stat_smoothers}
Auto %>% 
    ggplot(aes(x = mpg, y = horsepower)) +
    geom_point(alpha = 0.5) +
    stat_smooth(method = 'lm', se = FALSE) +
    stat_smooth(method = 'glm', formula = y ~ poly(x, 2), se = FALSE, color='orange') +
    stat_smooth(method = 'glm', formula = y ~ poly(x, 3), se = FALSE, color='red')

```

However if we have pre-computed data

```{r geom_bar_2}
# or just library(dplyr)
mpg_counted <- mpg %>% dplyr::count(class, name='count')
# notice the mapping
ggplot(mpg_counted) +
    geom_bar(aes(x=class, y=count), stat='identity')
```

```{r geom_col_2}
# or better
ggplot(mpg_counted) +
    geom_col(aes(x=class, y=count))
```

Use `after_stat()` to modify mapping from stats - i.e modify the values that
come out of the stats.

```{r after_stat}
ggplot(mpg) + geom_bar(aes(
    x=class,
    y=after_stat(100 * count / sum(count))
))
```

Some stats provide multiple calculate values 

```{r geom_density_1}
ggplot(mpg) +
    geom_density(aes(x=hwy))
```

```{r geom_density_2}
ggplot(mpg) +
    geom_density(aes(x=hwy, y=after_stat(scaled)))
```

You can add stat layers instead of geoms

```{r stat_count}
ggplot(mpg) +
    stat_count(aes(x=class))
```

```{r stat_summary}
# geom for stat_summary is point range
ggplot(mpg, aes(x = class, y = hwy)) +
    geom_jitter(width=0.2) +
    stat_summary(fun=mean, geom='point', colour='red', size=2)
```


## Scales

Naming scheme is `scale_<aesthetic>_<type>`. e.g `scale_x_continuous`


```{r scale_colour_brewer}
# type of column - continuous, factor, date
# use scale with colours that don't stand out
# qual means use a qualititive colour scale
ggplot(mpg) +
    geom_point(aes(x=displ, y=hwy, color=class)) +
    scale_colour_brewer(type='qual', name='Class of Vehicle')
# type palette='Set1'
```

Can control x and y

```{r scale_continuous}
diff(log10(c(10, 20, 30, 40)))
ggplot(mpg) +
    geom_point(aes(x=displ, y=hwy)) +
    scale_x_continuous(breaks=c(3, 5, 6)) +
    scale_y_continuous(trans='log10')
# short cut
# scale_y_log10()
```


Take a look at color brewer palettes

```{r inspect_colour_brewer}
RColorBrewer::display.brewer.all()
```

Bubble chart

```{r bubble_chart}
summary(as.factor(mpg$cyl))
# there is no 7
cylinders = as.integer(levels(as.factor(mpg$cyl)))
p<-ggplot(mpg) +
    geom_point(aes(x=displ, y=hwy, colour=class, size=cyl)) +
    scale_colour_brewer(type='qual')
p
p + scale_size(breaks=cylinders)
p + scale_radius(breaks=cylinders)
p + scale_size_area(breaks=cylinders)
```

```{r colour_and_size_legend}
p<-ggplot(mpg) +
    geom_point(aes(x=displ, y=hwy, colour=cty, size=cty))
p + scale_colour_continuous()
# merge colour and size
p + guides(colour='legend')
```

## Facets

Used to avoid overplotting. Facet grids are a way of doing pivot plots.

```{r facet_wrap_1}
p<-ggplot(mpg) +
    geom_point(aes(x=displ, y=hwy))
p+facet_wrap(~class)
```

```{r facet_grid_1}
p+facet_grid(year~drv)
```


Changing scales

```{r facet_scales}
p<-ggplot(mpg) +
    geom_point(aes(x=displ, y=hwy))
p + facet_wrap(~drv)
p + facet_wrap(~drv, scales='free')
p + facet_wrap(~drv, scales='free_y')
p + facet_wrap(~drv+year)
```

Changing the space occupied by each panel

```{r facet_panel_spacing}
p<-ggplot(mpg) +
    geom_bar(aes(y=manufacturer))
# all share the same scale. space wasted showing no data
p + facet_grid(class~.)
p + facet_grid(class~., space='free_y', scales='free_y')
```


## Coordinates

How should x and y be interpreted?

```{r polar_coordinates}
p<-ggplot(mpg) +
    geom_bar(aes(x=class))
# x->angle, y->radius
p + coord_polar()
# change what gets mapped to angle
p + coord_polar(theta='y') + expand_limits(y=70)
```


### Zooming

```{r zooming}
p<-ggplot(mpg) +
    geom_bar(aes(x=class))
# this removes bars
p + scale_y_continuous(limits=c(0, 40))
# this is better
p + coord_cartesian(ylim=c(0, 40))
```

Just as you can set limits in coordinates as well as scales, so to can you do transformations in coordinates.

Scale transforms the input values prior to calculating the breaks

```{r scale_transforming}
p<-ggplot(mpg) +
    geom_point(aes(x=hwy, y=displ))
p
p + scale_y_continuous(trans='log10')
p + scale_y_continuous(trans='log')
# breaks have already been calculated. we are just stretching the fabric
p + scale_y_continuous(labels=c(1:7), breaks=c(1:7)) + coord_trans(y='log')
```


## Maps

```{r world_map}
world <- sf::st_as_sf(maps::map('world', plot=FALSE, fill=TRUE))
world <- sf::st_wrap_dateline(world, options=c("WRAPDATELINE=YES", "DATELINEOFFSET=180"), quiet=TRUE)

p<-ggplot(world) +
    geom_sf()
p + coord_sf(crs='+proj=moll')
```

## Themes

Not related to data

```{r custom_theme_initial_plot}
p<-ggplot(mpg) +
    geom_bar(aes(y=class)) +
    facet_wrap(~year) +
    labs(
        title='Number of car models per class',
        caption='source: http//fueleconomy.gov',
        x=NULL,
        y=NULL
    )

p
```

```{r minimal_theme}
p + scale_x_continuous(expand = c(0, NA))

p + theme_minimal()
```

```{r custom_theme}
new_theme = theme(
    text = element_text('Avenir Next Condensed'),
    strip.text = element_text(face='bold', hjust=0),
    plot.caption = element_text(face='italic'),
    panel.grid.major = element_line('white', size=0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.ontop = TRUE
)

p + theme_minimal() + new_theme
```


