---
title: "GGplot2 Workshop part 2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(patchwork)
library(gganimate)
library(ggrepel)
library(ggforce)
```


## Plot Composition

We use patchwork

```{r patchwork_1}
p1<-ggplot(msleep) +
    geom_boxplot(aes(x=sleep_total, y=vore, fill=vore))
p2<-ggplot(msleep) +
    geom_bar(aes(y=vore, fill=vore))
p3<-ggplot(msleep) +
    geom_point(aes(x=bodywt, y=sleep_total, colour=vore)) +
    scale_x_log10()

# | combines. / is the stack operator
p_all <- (p1 | p2)/p3
p_all + plot_layout(guides='collect')
p_all & theme(
    legend.position='none'
)
p_all + plot_annotation(
    title='Mammalian sleep patterns',
    tag_levels = 'A'
)
```

We can customize how much space each plot takes

```{r patchwork_space}
p<-ggplot(mtcars) +
    geom_point(aes(x=disp, y=mpg))
plots<-p + p + p
# relative sizes
plots + plot_layout(widths=c(1,2,1))
# first plot gets 5 cm, the other two split the remaining space between them
plots + plot_layout(width=unit(c(6, 1, 1), c('cm', 'null', 'null')))
```

More on the `&` operator

```{r patchwork_composition}
p1 <- ggplot(mtcars[mtcars$gear == 3,]) +
    geom_point(aes(x=disp, y=mpg))
p2 <- ggplot(mtcars[mtcars$gear == 4,]) +
    geom_point(aes(x=disp, y=mpg))
plots <- p1 + p2
# note that the axis values are not aligned
plots
plots & scale_y_continuous(limits=c(10,40), name = NULL)
```


Layouts can be specified by design

```{r patchwork_layouts}
p1<-ggplot(mtcars) +
    geom_point(aes(x=disp, y=mpg))
p2<-ggplot(mtcars) +
    geom_bar(aes(x=factor(gear)))
p3<-ggplot(mtcars) +
    geom_boxplot(aes(x=factor(gear), y=mpg))
# 3x3 grid (plot, plot, blank)(blank, plot, plot)(plot, blank, blank)
layout<-'AA#\n#BB\nC##'
plots <- p1 + p2 + p3
plots + plot_layout(design=layout)
```

## Animation

Continuous transitions

```{r continouos_transition}
# start with something static
p<-ggplot(economics) +
    geom_line(aes(x=date, y=unemploy)) +
    transition_reveal(along=date)
# now add a transition that reveals the geom (line) along the date axis
animate(p, renderer = av_renderer())

```

Discrete transitions

```{r discrete_transition}
# place holder. uses glue package. every thing in curly brackets is evaluated as R code
# closest_state will be the year in this case
# i.e the name of the state that is closest to the current frame
p<-ggplot(mpg) +
    geom_bar(aes(x=factor(cyl))) +
    labs(title='Number of cars in {closest_state} by number of cylinders')

# enter and exit because new graphic data comes in (5 cylinders in this case)
p + transition_states(states=year) +
    enter_grow() +
    exit_fade()

```

The animation below will animate points showing cars with different cylinders

```{r car_cylinders}
p<-ggplot(mpg) +
    geom_point(aes(x=displ, y=hwy)) +
    ggtitle("Cars with {closest_state} cylinders")
# it looks like the cars are changing cylinders
p + transition_states(factor(cyl))

```

use the group aesthetic to match observations between states


```{r car_cylinders_with_groups}
# give each row its own unique group
row_ids = seq_len(nrow(mpg))
p<-ggplot(mpg) +
    geom_point(aes(x=displ, y=hwy, group=row_ids)) +
    ggtitle("Cars with {closest_state} cylinders")
# it looks like the cars are changing cylinders
p + transition_states(factor(cyl))
# if you say colour=factor(cyl) you'll get the same effect
```


Easing refers to the speed at which values change during an animation

```{r easing}
# cty and hwy move from being columns to rows
mpg2<-tidyr::pivot_longer(mpg, c(cty, hwy))
p<-ggplot(mpg2) +
    geom_point(aes(x=displ, y=value)) +
    ggtitle("{if (closest_state == 'city') 'Efficiency in city' else 'Efficiency on highway'}")
# from constant speed from y aesthetic
p + transition_states(name) +
    ease_aes(y='cubic-in-out')
```

View the documentation by typing `?animate`


## Annotations

Overplotting

```{r overplotting_example}
ggplot(mtcars, aes(x=disp, y=mpg)) +
    geom_point() +
    geom_text(aes(label=row.names(mtcars)))
```

Now with `ggrepel`

```{r ggrepel_example, fig.width=10, fig.height=8}
labels = row.names(mtcars)
ggplot(mtcars, aes(x=disp, y=mpg)) +
    geom_point(size = 2, alpha = .8) +
    geom_text_repel(aes(label = labels), color = 'steelblue')

```

highlighting with `ggforce`

```{r ggforce_highlighting}
p<-ggplot(mtcars, aes(x=disp, y=mpg)) +
    geom_point() +
    coord_cartesian(xlim=c(0, 500), ylim=c(0,40))
p + geom_mark_ellipse(aes(
        filter=gear==4,
        label='4 gear cars',
        description='Cars with fewer gears tend to both have higher yield and lower displacement'
    ))

p + geom_mark_hull(aes(fill=factor(gear), label=gear))
```


Hiding guides

```{r hiding_guides}
ggplot(mtcars, aes(x=disp, y=mpg, colour=factor(gear))) +
    geom_point() +
    guides(colour=guide_none())
```



