---
title: "Creating Heatmaps"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE
)
options(digits=3)
library(tidyverse)
library(gplots)
library(pheatmap)
```

## Using base `heatmap` with `mtcars`

Heatmaps convert a 2 dimensional numeric matrix to a colour


```{r cars_matrix}
mtcars_mat <- data.matrix(mtcars)
head(mtcars_mat)
```

```{r cars_heatmap_1}
heatmap(mtcars_mat)
```

too much. look at the scale argument. The default is row (i.e individual cars). We want the column direction. we want the highs and lows in our variables.

```{r cars_heatmap_2}
heatmap(mtcars_mat, scale = 'column')
```

colours - look at `cm.colors`

```{r cars_heatmap_3}
heatmap(mtcars_mat, scale = 'column', col = cm.colors(n=nrow(mtcars_mat)))
```
You could also take a look at `hcl.pals()`

Remove the dendogram

```{r cars_heatmap_4}
heatmap(mtcars_mat, scale='column', Colv = NA)
```
```{r cars_heatmap_cleanup}
rm(mtcars_mat)
```


## Using gplots


```{r load_heatmap_data}
protein_treatments_df <- read_csv(
    '../data/protein_treatments.csv',
    col_types = cols(
        Row_ID = col_integer()
    )
)
```


```{r heatmap_data_etl}
protein_treatments_df <- protein_treatments_df %>% 
    # remove the spaces from the column names
    rename_with(~str_replace(.x, ' ', '_')) %>% 
    # and we need rownames
    column_to_rownames(var='Row_ID')
```


```{r heatmap_data_matrix}
# remove the t-test variables
protein_treatments_mat <- protein_treatments_df %>% 
    select(-T_test) %>% 
    data.matrix()
head(protein_treatments_mat)
```

scale in the row direction - how each protein has reacted to each treatment: `heatmap.2` does the clustering before the scaling


```{r heatmap_2_initial, fig.height=10, fig.width=10}
gplots::heatmap.2(protein_treatments_mat, scale='row',  cexCol = 0.8)
```

It helps to

1. scale the data before visualisation (`?scale`)
2. transpose the matrix so that the rows are scaled (`?t`)


```{r data_scaled}
# transpose, scale then transpose back
data_scaled <- t(scale(t(protein_treatments_mat)))
```

protein responses are now comprable to each other

```{r heatmap_2_scaled, fig.height=12, fig.width=10}
gplots::heatmap.2(protein_treatments_mat, scale='row', cexCol = 0.8)
```



```{r heatmap_2_palette, fig.height=12, fig.width=10}
my_palette <- grDevices::colorRampPalette(c('blue', 'white', 'red'))
heatmap.2(
    data_scaled,
    col = my_palette(n=20),
    trace = 'none',
    dendrogram = 'row',
    main = 'Protein response',
    margins = c(6, 4),
    cexRow = 0.8,
    cexCol = 0.8
)
```

- top right: proteins that were upregulated by the treatments
- bottom left: regulated by the treatments


```{r heatmap_2_cleanup}
dev.off()
rm(protein_treatments_df, protein_treatments_mat, my_palette, data_scaled)
```

## Using `pheatmap`


```{r pheatmap_data}
data_m <- matrix(rnorm(25), 5, 5)
colnames(data_m) <- paste0('Treatment', 1:5)
rownames(data_m) <- paste0('Gene', 1:5)
```


```{r pheatmap_plot}
pheatmap(
    data_m,
    title='Pretty heatmap',
    cellwidth = 50,
    cellheight = 30,
    fontsize = 11,
    display_numbers = TRUE,
    scale = 'row'
    # filename='plots/pheatmap.pdf'
)
```


```{r pheatmap_cleanup}
rm(data_m)
```


## Using `ggplot2`

We will be using the `esoph` dataset (`?esoph`)


```{r esoph_data}
esoph_sub <- esoph %>%
    filter(agegp == '55-64') %>% 
    mutate(frequency = ncases / (ncases + ncontrols))
```

how frequent are cases in this population?


```{r esoph_plot}
# 2 categorical varaiables to create the cells, and a numerical value to fill them
esoph_sub %>% 
    ggplot(aes(x = alcgp, y = tobgp, fill = frequency)) +
    geom_tile(color = 'white') +
    scale_fill_gradient(low = 'white', high = 'steelblue') +
    theme_minimal() +
    labs(fill = 'Cancer frequency', x = 'Alcohol consumption', y = 'Tobacco consumption')
```



