---
title: "Conditioning"
output:
    html_document: default
---


```{r setup, echo=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    error = T,
    message = F,
    warning = F
)
options(digits = 3)
```

What does it mean when we say that a system is *ill-conditioned*?

An ill-conditioned system is one whose determinants are close to zero.

Lets say we have the following system:

$$
x_1 + x_2 = 2 \\
2 x_2 + 2 x_2 = 4
$$
This can be written as a linear system with a square matrix:

$$
\left[ \begin{array}{cc}
1 & 1 \\
2 & 2
\end{array}\right]
\left[\begin{array}{c}
x_1 \\
x_2
\end{array}\right] = 
\left[ \begin{array}{c}
2 \\
4
\end{array}\right]
$$

Which has no solution, becuase the second equation is the same as the first

```{r example_3_not_solvable}
A = cbind(c(1, 2), c(1, 2))
det(A)
```
Now consider this system:

$$
\left[ \begin{array}{cc}
2 & 1 \\
4 & 2.001
\end{array}\right]
\left[\begin{array}{c}
x_1 \\
x_2
\end{array}\right] = 
\left[ \begin{array}{c}
3 \\
6
\end{array}\right]
$$

This is solvable, but is *numerically unstable*. Note how small the determinant is and how large the inverse due to the division by the extremely small determinant.

```{r example_3_unstable}
A = cbind(c(2, 4), c(1, 2.001))
b = c(3, 6)
AI <- MASS::ginv(A)
det(A)
AI
```

The solution to this system is as follows:

```{r example_3_solve1}
solve(A, b)
```

But change $\vec{b}$ just a little and make $\vec{b} = \left[ \begin{array}{c}3 \\ 6.01 \end{array} \right]$ and the solution is drastically different:

```{r example_3_solve2}
solve(A, c(3, 6.01))
```

This is inherently unstable and is a consequence of the near singularity of the matrix `A`.

Such linear systems are called ill-conditioned.

