---
title: "Quasi Separation"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(digits = 4)
```


```{r load_car_data}
car_data <- read_csv(
    '../data/car_ratings.csv',
    col_types = cols(
        buying = col_factor(),
        maint = col_factor(),
        doors = col_factor(),
        persons = col_factor(),
        lug_boot = col_factor(),
        rating = col_factor()
    )
) %>% rename(
    car_price = buying,
    maint_price = maint
) %>% mutate(
    fail = (rating == 'unacc')
)
```

```{r car_data_response_variable}
prop.table(xtabs(~fail, data=car_data))
```



```{r car_train_test_split}
# the carat way
set.seed(24351)
car_split <- caret::createDataPartition(
    car_data$fail,
    p = 0.7,
    list = FALSE
)
car_train <- car_data %>%
    slice(car_split)

car_test <- car_data %>%
    slice(-car_split)

nrow(car_data)
nrow(car_train)
nrow(car_test)

prop.table(xtabs(~fail, data=car_train))
prop.table(xtabs(~fail, data=car_test))

```

```{r save_car_train_and_test}
write_rds(car_data, '../data/car_rating.rds')
```


### Fit the regression model

```{r fit_car_model}
(fmla <- wrapr::mk_formula(
    'fail',
    setdiff(colnames(car_data), c('rating', 'fail'))
))

car_model <- glm(fmla, data = car_train, family = binomial)
```

This warning indicates that the problem is quasi-separable: some set of variables perfectly predicts a subset of the data.

In fact, this problem is simple enough that you can easily determine that a safety rating of low perfectly predicts that a car will fail the review.

```{r failure_counts_for_low_safety_cars}
car_train %>%
    filter(safety == 'low') %>%
    count(fail)
```

However, even cars with higher safety ratings can get ratings of unacceptable, so the safety variable only predicts a subset of the data.

```{r}
xtabs(~rating+safety, data = car_train)
xtabs(~fail+safety, data = car_train)
```



```{r}
summary(car_model)
```

The variables `safetylow`, `persons4`, and `personsmore` all have unusually high magnitudes and very high standard errors. As mentioned earlier, `safetylow` always corresponds to an unacceptable rating, so `safetylow` is a strong indicator of failing the review. However, larger cars (cars that hold more people) are not always going to pass the review.

It’s possible that the algorithm has observed that larger cars tend to be safer (get a safety rating better than safetylow), and so it is using the persons4 and personsmore variables to cancel out the overly high coefficient from safetylow.

In addition, you can see that the number of Fisher scoring iterations is unusually high; the algorithm did not converge.

This problem is fairly simple, so the model may predict acceptably well on the test set; however, in general, when you see evidence that `glm()` did not converge, you should not trust the model.

```{r}
coef_df <- broom::tidy(car_model)
coef_df %>%
    slice(-1) %>%
    ggplot(aes(x = term, y = estimate)) +
    geom_pointrange(aes(ymin = 0, ymax = estimate)) +
    labs(x = NULL) +
    ggtitle('Coefficients of logistic regression model') +
    coord_flip()

```

In the plot, coefficients that point to the right are positively correlated with failing the review, and coefficients that point to the left are negatively correlated with failure.


```{r logistic_model_evaluation_routine}
# print confutions matrix, accuracy and deviance
confmat <- function(dframe, predvar) {
    cmat <- table(
        truth = ifelse(dframe$fail, "unacceptable", "passed"),
        prediction = ifelse(dframe[[predvar]] > 0.5, "unacceptable", "passed")
    )
    accuracy <- sum(diag(cmat)) / sum(cmat)
    deviance <- sigr::calcDeviance(dframe[[predvar]], dframe$fail)

    list(
        confusion_matrix = cmat,
        accuracy = accuracy,
        deviance = deviance
    )
}

```

```{r evaluate_logistic_model}
car_test <- car_test %>%
    modelr::add_predictions(car_model, type = 'response', var = 'pred_glm')
confmat(car_test, 'pred_glm')
```


In this case, the model seems to be good. However, you cannot always trust non-converged models, or models with needlessly large coefficients.
