---
title: "Roots of ARMA(p, q) Models"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=3)
source('../lib/eng_maxima.R')
knitr::knit_engines$set(maxima = eng_maxima)
```


## Simple Example

Find the root to determine if a stochastic process is stationarity, invertible or causal.

General form:

$$
\phi(B) X_t = \theta(B)Z_t
$$

Where $Z_t$ is the noise and $X_t$ is our time series (the stochastic process) and `B` is the shift operator:

$$
B^j X_t = X_{t - j}
$$

For example $B X_t = X_{t-1}$ and $B^2X_t=X_{t-2}$. The same follows for $Z_t$

We can now write ARMA models as polynomials, whose roots we can find. These roots determine the following properties:

### Stationarity

&phi; roots are **not** on the unit circle

### Causality

&phi; roots are **outside** the unit circle

### Invertibility

&theta; roots are **outside** unit circle


## Example 1: AR(1) model


First state the model in standard (AR) form:

$$
X_t = 0.9 X_{t-1} + Z_t
$$

Lets take a quick look at it

```{r}
arma_1 <- arima.sim(
    model = list(
        order = c(1, 0, 0),
        ar = 0.9,
        sd = sqrt(2)
    ),
    n = 100
)

forecast::tsdisplay(
    arma_1,
    main = 'ARMA(1, 0) Model with ar1=0.9'
)
```

Now write the stochastic process as a polynomial using the backward shift operator:

First put the expressions for $X_t$ and $Z_t$ on the left and right sides of the expression.

$$
X_t - 0.9 X_{t-1} = Z_t
$$

Next write it in terms of backward shift operators:

$$
(1 - 0.9B)X_t = Z_t
$$

Distribute $X_t$ to get back our previous expression. Now write the characteristic functions for &phi; and &theta;

$$
\phi(z) = 1-0.9z \\
\theta(z) = 1
$$


Now we can find the roots of `z`. There are no roots for &theta;, so the time series is **invertible**.

Next we solve for &phi;

```{maxima maxima_1, echo = FALSE}
fz: 1 - 0.9 * z = 0;
solve(fz, z);
```

And we get `z = 10/9` or `1.1`. Since its greater than 1 we have **causality**. Since its not equal to 1 we also have **stationarity**.

## Example 2: ARMA(2, 1)

$$
X_t = .75 X_{t-1} - .5625 X_{t-2} + Z_t + 1.25 Z_{t-1}
$$

Visualize it

```{r arma_21}
arma_21 <- arima.sim(
    model = list(
        order = c(2, 0, 1),
        ar = c(.75, -0.5625),
        ma = 1.25,
        sd = sqrt(2)
    ),
    n = 100
)

forecast::tsdisplay(
    arma_21,
    main = 'ARMA(2, 1) Model with ar(.75, -.5625) ma(1.25)'
)
```


Collect terms

$$
X_t - .75 X_{t-1} + .5625 X_{t-2} = Z_t + 1.25 Z_{t-1}
$$

State in terms of backshift operations

$$
(1 - .75B + .5625B^2) = (1 + 1.25)B
$$


Next derive the characteristic polynomials

$$
\phi(z) = 1 - .75z + .5625 z^2 = 0\\
\theta(z) = 1 + 1.25z = 0
$$

Finally, solve for `z` and test

```{maxima maxima_2, echo = FALSE}
tz: 1 + 1.25 * z = 0;
pz: 1 - .75 * z + .5625*z^2 = 0;
tsol: solution(tz, z);
psol: solution(pz, z);
psol: rectform(psol);
abs(psol);
```

So we get $\theta(z) = -4/5$ or $\theta(z) = -0.8$.

This is less then -1 so its inside the unit circle an so the series is **not** invertible.

$\phi(z) = 4/3$. This is not equal to one, so we have *stationarity* and because its greater than one it is also *causal*.


