---
title: "Association Rules with the Movielens dataset"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    fig.width = 10,
    fig.height = 8,
    paged.print=FALSE
)
options(digits=5)
library(tidyverse)
library(arules)
library(zeallot)
```


```{r movie_data, paged.print=TRUE}
movies <- read_csv(
    '../data/movielens.csv',
    col_types = cols(
        movieId = col_integer(),
        title = col_character(),
        genres = col_character()
    )
)
movies %>% 
    head()

```

```{r movies_extract_title_and_year}
# separate title and year
title_year <- '(.+) \\(([0-9]{4})\\)'

sample <- movies %>% 
    sample_n(10) %>% 
    select(title)

str_match(sample$title, title_year)

# 100x3 matrix
# movies %>% 
#     sample_n(5) %>% 
#     .$title %>% str_match(title_year)

sample %>% 
    extract(title, c('name', 'year'), title_year, remove=F, convert=T)

movies <- movies %>% extract(
    title,
    c('name', 'year'),
    title_year,
    remove = F,
    convert=T
)


movies <- movies %>% 
    filter(!is.na(name), year>=1900) %>% 
    mutate(name=str_trim(name)) %>% 
    select(-title) %>% 
    rename(title=name)

rm(title_year)
```

```{r movies_extract_alternative_title_practice}
# practice first
movies %>% 
    sample_n(100) %>% 
    pull(title)

# match everything inside of brackets thats not acually an end bracket

bracketed_term = '\\(([^)]+)\\)'

samp <-  tibble(
    movie = c('Days of Grace (Días de gracia)')
)
remove_bracketed_term <- function(values) {
    values %>% str_replace(bracketed_term, '') %>% str_trim()
}

samp %>% 
    extract(movie, bracketed_term, into='alt_title_1', remove=F) %>% 
    mutate(movie=remove_bracketed_term(movie))


movies %>% 
    sample_n(100) %>% 
    extract(title, bracketed_term, into='alt_title_1', remove=F) %>% 
    mutate(
        title = ifelse(
            !is.na(alt_title_1),
            remove_bracketed_term(title),
            title
        )
    ) %>% 
    filter(!is.na(alt_title_1)) %>% 
    select(title, alt_title_1)



rm(samp)
write_rds(movies, '../staging/movies.rds')
movies <- read_rds('../staging/movies.rds')
```

```{r movies_extract_alternative_title_practice}
movies <- movies %>% 
    extract(title, bracketed_term, into='alt_title_1', remove=F) %>% 
    mutate(
        title = ifelse(
            !is.na(alt_title_1),
            remove_bracketed_term(title),
            title
        )
    )

movies %>% 
    sample_n(200) %>% 
    filter(!is.na(alt_title_1)) %>% 
    arrange(title) %>% 
    select(title, alt_title_1)

# go again
movies <- movies %>% 
    extract(title, bracketed_term, into='alt_title_2', remove=F) %>% 
    mutate(
        title = ifelse(!is.na(alt_title_2), remove_bracketed_term(title), title)
    )

movies %>% 
    filter(str_detect(title, bracketed_term)) %>% 
    select(title, alt_title_1, alt_title_2)

# go again
movies <- movies %>% 
    extract(title, bracketed_term, into='alt_title_3', remove=F) %>% 
    mutate(
        title = ifelse(!is.na(alt_title_3), remove_bracketed_term(title), title)
    )

movies %>% 
    filter(str_detect(title, bracketed_term)) %>% 
    select(title, starts_with('alt'))

# one last time
movies <- movies %>% 
    extract(title, bracketed_term, into='alt_title_4', remove=F) %>% 
    mutate(
        title = ifelse(!is.na(alt_title_4), remove_bracketed_term(title), title)
    )

movies %>% 
    filter(str_detect(title, bracketed_term)) %>% 
    select(title, starts_with('alt'))

movies <- movies %>% 
    mutate(title=str_trim(title))
```


```{r movies_view_subtitles}
movies %>% 
    filter(str_detect(title, ': ?')) %>% 
    select(title)
```

Splitting these will help us find movies that are sequels of one another


```{r subtitle_exploration}
title_subtitle_re <- '([^:]+): ?(.+)'
subtitle_re <- ': ?(.+)'
samp <- tibble(
    title = c('Ace Ventura: When Nature Calls')
)
samp %>% 
    extract(title, title_subtitle_re, into=c('title', 'subtitle'))

samp %>% 
    extract(title, subtitle_re, into=c('subtitle'), remove=F)

movies %>% 
    filter(str_detect(title, ': ?')) %>% 
    extract(title, subtitle_re, into='subtitle', remove=F) %>% 
    mutate(
        title = ifelse(
            !is.na(subtitle),
            str_replace(title, subtitle_re, ''),
            title
        )
    ) %>% 
    modify_at(vars(title, subtitle), str_trim)


```


```{r subtitle_extraction}
movies <- movies %>% 
    extract(title, subtitle_re, into='subtitle', remove=F) %>% 
    mutate(
        title = ifelse(
            !is.na(subtitle),
            str_replace(title, subtitle_re, ''),
            title
        )
    ) %>% 
    modify_at(vars(title, subtitle), str_trim)

write_rds(movies, '../staging/movies.rds')
```


```{r movie_prefixes_exploration}
prefix_re <- ', ([^ ]{1,5})$'
end_brackets <- '[:)]+$'

movies %>% 
    filter(str_detect(title, end_brackets)) %>% 
    select(title, subtitle)

# remove stay brackets at the end
movies %>% 
    filter(str_detect(title, end_brackets)) %>% 
    mutate(title = str_replace(title, end_brackets, ''))


movies <- movies %>% 
    mutate(title = str_trim(str_replace(title, end_brackets, '')))

movies %>% 
    filter(str_detect(title, prefix_re)) %>% 
    select(title)

prefixes <- str_match(movies$title, prefix_re)[,2]
unique(prefixes)
movies %>% 
    filter(str_detect(title, ', (De|Da)$')) %>% 
    select(title)
movies %>% 
    filter(str_detect(title, prefix_re)) %>% 
    filter(!str_detect(title, ', (A|The)')) %>% 
    select(title)

prefixes <- ", (The|the|A|An|But|Da|De|Die|Das|Der|Los|La|Le|Les|Las|En|Une|L|L'|I|Il|El|O)$"
movies %>% 
    filter(str_detect(title, prefixes)) %>% 
    extract(title, prefixes, into='prefix', remove=F) %>% 
    filter(!prefix %in% c('The', 'A')) %>% 
    arrange(prefix) %>% 
    select(title, prefix)

```

```{r movie_prefixes_extraction}
movies <- movies %>% 
    extract(title, prefixes, into='prefix', remove=F) %>% 
    mutate(
        title = ifelse(
            !is.na(prefix),
            str_trim(str_replace(title, prefixes, '')),
            title
        )
    )
write_rds(movies, '../staging/movies.rds')
movies = read_rds('../staging/movies.rds')
```

Check for odd titles

```{r check_title_lengths}
movies %>% 
    mutate(title_length = str_length(title)) %>% 
    filter(title_length > 60) %>% 
    arrange(desc(title_length)) %>% 
    select(title, year)
```

Looks like we have another pattern for subtitles

```{r more_subtitles, paged.print=TRUE}
subtitle_re <- ' - (.+)$'
samp <- tibble(
    title = c('Kino-pravda no. 13 - Yesterday, Today, Tomorrow. A Film Poem Dedicated to the October Revolution')
)
samp %>% 
    extract(title, subtitle_re, into='subtitle', remove=F)

movies %>% 
    filter(str_detect(title, subtitle_re)) %>% 
    extract(title, subtitle_re, into='subtitle2', remove=F) %>% 
    select(title, subtitle2)

movies <- movies %>% 
    extract(title, subtitle_re, into='subtitle2', remove=F) %>% 
    mutate(
        title = ifelse(
            !is.na(subtitle2),
            str_trim(str_replace(title, subtitle_re, '')),
            title
        )
    ) %>% 
    mutate(subtitle = ifelse(
        !is.na(subtitle2),
        subtitle2,
        subtitle
    )) %>%
    select(-subtitle2)
    
```

```{r movie_titles_with_or}
subtitle_re <- ', or (.{10,})$'
movies %>% 
    filter(str_detect(title, subtitle_re)) %>% 
    select(title)
```


```{r extract_or_subtitles}
movies %>% 
    filter(str_detect(title, subtitle_re)) %>% 
    extract(title, subtitle_re, into='subtitle2', remove=F) %>% 
    select(title, subtitle2)

movies <- movies %>% 
    extract(title, subtitle_re, into='subtitle2', remove=F) %>% 
    mutate(
        title = ifelse(
            !is.na(subtitle2),
            str_trim(str_replace(title, subtitle_re, '')),
            title
        )
    ) %>% 
    mutate(subtitle = ifelse(!is.na(subtitle2), subtitle2, subtitle)) %>%
    select(-subtitle2)
```


```{r check_title_lengths_again}
movies %>% 
    mutate(title_length = str_length(title)) %>% 
    filter(title_length > 60) %>% 
    arrange(desc(title_length)) %>% 
    select(title)
```



```{r reorder_columns_and_save}
movies <- movies %>% 
    select(movieId, prefix, title, subtitle, year, genres, alt_title_1, alt_title_2, alt_title_3, alt_title_4)

movies %>% write_rds('../staging/movies.rds')
# clean up
rm(samp, bracketed_term, end_brackets, prefix_re, prefixes, subtitle_re, title_subtitle_re)
```

ETL Done

```{r load_movies_post_title_etl}
movies <- read_rds('../staging/movies.rds')
# we don't need all the alternative titles in this analysis
movies <- movies %>% 
    select(!starts_with('alt'))
```



```{r movie_genres}

all_genres <- unique(unlist(str_split(movies$genres, '\\|')))
all_genres
#  movies without genres
movies %>% 
    filter(str_detect(genres, '(no genres listed)')) %>% 
    nrow()
```

Create indicator columns for each genre.

```{r}
movie_genres <- movies %>% 
    # we just need the movie id and the genres column
    select(movieId, genres) %>% 
    tidytext::unnest_regex(genre, genres, '\\|') %>%
    filter(!genre %in% c('imax', '(no genres listed)'))

movie_genres %>% 
    .$genre %>% 
    fct_count(sort = TRUE)
```

Look at the distribution by genre

```{r movies_by_genre, fig.width = 8, fig.height = 7}
movie_genres %>% 
    ggplot(aes(x = fct_rev(fct_infreq(genre)))) +
    geom_bar(fill='gray80', colour='gray50') +
    labs(x = NULL, y = NULL) +
    coord_flip() +
    theme(legend.position = 'none')
```


Another way of doing it


```{r}
samp <- movies %>% 
    sample_n(100)
all_genres = all_genres[! all_genres %in% c("IMAX","(no genres listed)")]

for(genre in all_genres) {
    samp[str_c("genre_", genre)] = ifelse((str_detect(samp$genres,genre) | str_detect(samp$genres, "no genres")), 1, 0)
}
samp$genres <- NULL
rm(samp, all_genres)

```

Convert genres into features and add to our movies

```{r extract_genre_features}
movies_wide <- movie_genres %>% 
    mutate(val = 1) %>% 
    pivot_wider(names_from = genre, values_from = val, values_fill = 0) %>%
    modify_if(is.double, as.integer)

movies <- movies %>% 
    inner_join(movies_wide) %>% 
    select(-genres)

write_rds(movie_genres, '../staging/movie_genres.rds')
write_rds(movies, '../staging/movies_complete.rds')
rm(movie_genres, movies_wide, all_genres)
```

This completes the feature engineering of movies

```{r load_completed}
movies <- read_rds('../staging/movies_complete.rds')
```



Explore the number of movies for each year in the dataset

```{r movies_by_year, fig.width = 12, fig.height = 7}
movies %>% 
    ggplot(aes(year)) +
    geom_bar(colour = 'gray50', fill = 'gray80') +
    scale_x_continuous(
        breaks = seq(1900, 2020, 10)
    ) +
    ggtitle('Number of movies by year')
```



Load the ratings


```{r load_movie_ratings}
ratings <- read_csv(
    '../data/movie_ratings.csv.gz',
    col_types = cols(
        userId = col_integer(),
        movieId = col_integer(),
        rating = col_skip(),
        timestamp = col_skip()
    )
)

```

```{r combine_movies_and_ratings}
# discard ratings whose movie id has been discarded
ratings <- ratings %>% 
    semi_join(movies)
# how many ratings
dim(ratings)
```


We construct User-Item matrix with binary values; 0 - a movie isn’t seen by a user, and 1 - it is seen. The package uses a sparse matrix object, `transactions`, to represent the User-Item matrix.

There will be a row for each unique `userId` and indicator columns for each `movieId`

This prevents us from consuming all available RAM as most elements in the matrix will be zero.

```{r split_function}
samp <- ratings[1:100,]
# see what split does
base::split(samp[,'movieId'], samp[,'userId'])
```

```{r test_user_item_matrix}
user_item_matrix <- as(split(samp[,"movieId"], samp[,"userId"]), "transactions")
dim(user_item_matrix)
```



```{r user_item_matrix}
user_item_matrix <- as(split(ratings[,"movieId"], ratings[,"userId"]), "transactions")
dim(user_item_matrix)
dim(ratings)
# the number of unique items in each column
apply(ratings, 2, function(col) length(unique(col)))
c(num_users, num_movies) %<-% dim(user_item_matrix)

# dont' need this anymore
rm(ratings, samp)
```

```{r user_item_matrix_summary}
user_item_matrix
```


Next, we mine for a frequent pair of movies that raters watched. We hypothesize that if movie A and B are frequently viewed together, there should be some underlying relationships between them that incite viewer’s curiosity.

We can use such finding to recommend movie B to a user if he/she already saw A (or vice versa).

We set the support threshold to `0.001`, i.e the pair is watched together by at least `{r round(num_users * 0.001)}`.

We set the minimum confidence (the likelihood that is a user watched movie A they also watch movie B) to 70%.


```{r build_the_model, paged.print=FALSE}

rule_params <- list(
    supp = 0.001,
    conf = 0.7,
    maxlen = 2
)
assoc_rules <- apriori(user_item_matrix, parameter = rule_params)
summary(assoc_rules)
# don't need this anymore
rm(user_item_matrix)
```

Given so many rules, we filter only those that have a lift greater than 4.25


```{r assoc_rules_lift}
rules_df <- as(assoc_rules, 'data.frame')
quantile(rules_df$lift, probs=c(0.25, 0.50, 0.75))

rules_df <- rules_df %>% 
    filter(lift >= 4.25) %>% 
    modify_if(is.double, round, digits=3)
rm(assoc_rules, rule_params)

rules_df %>% 
    sample_n(10)
```

This is a terrible way to extract the movie ids

```{r extract_movie_ids}

# convert {705} -> {95} to 705,95
samp <- rules_df %>% 
    sample_n(1000)
rules = sapply(samp$rules,function(x){
    x = gsub("[\\{\\}]", "", regmatches(x, gregexpr("\\{.*\\}", x))[[1]])
    x = gsub("=>",",",x)
    x = str_replace_all(x," ","")
    return( x )
})

head(rules)

# remove column names from vector
rules = as.character(rules)
head(rules)
rules = str_split(rules,",")
head(rules)

samp$lhs_movie = sapply(rules, "[[", 1)
samp$rhs_movie = sapply(rules , "[[", 2)

samp$rules = NULL
# and you still have to convert them to integers
samps$lhs_movie = as.numeric(samp$lhs_movie)
samps$rhs_movie = as.numeric(samp$rhs_movie)
samp %>% 
    head()
rm(samp)
rm(rules)
gc()

```


```{r extract_movie_ids_the_tidy_way}
movie_id_pattern = '\\{([0-9]+)\\} => \\{([0-9]+)\\}'

tibble(movie=c('{567} => {671}')) %>%
    extract(movie, movie_id_pattern, into=c('lhs_movie', 'rhs_movie'), convert = T)

rules_df <- rules_df %>% 
    # this will remove the rules column
    extract(rules, movie_id_pattern, into=c('lhs_movie', 'rhs_movie'), convert = T)
rm(movie_id_pattern)
head(rules_df)
write_rds(rules_df, '../staging/rules_df.rds')
rules_df <- read_rds('../staging/rules_df.rds')
```


```{r add_movie_titles, paged.print=TRUE}
n_cols <- ncol(movies)
start_col <- ncol(rules_df) + 1
rules_df %>% left_join(movies, by=c('lhs_movie' = 'movieId')) %>% 
    head()
start_col:(n_cols - 2)
# titles and years for left hand side
# rules_df <- 
rules_df <- rules_df %>% left_join(movies, by=c('lhs_movie' = 'movieId')) %>% 
    rename_at(start_col:(start_col + n_cols - 2), ~str_c('left_', .))

colnames(rules_df)
start_col <- ncol(rules_df) + 1
# and the right hand side
rules_df <- rules_df %>% left_join(movies, by=c('rhs_movie' = 'movieId')) %>% 
    rename_at(start_col:(start_col + n_cols - 2), ~str_c('right_', .))

colnames(rules_df)
rm(n_cols, start_col)
write_rds(rules_df, '../staging/rules_df.rds')
```


```{r top_rules_with_highest_lift}
rules_df %>% 
    arrange(desc(lift)) %>% 
    select(left_title, left_year, right_title, right_year, lift, count) %>% 
    head(n=50)
```

TODO: identiy sequels


Lets look at modern movies that led users to view the older film.

```{r old_new}
rules_df %>% 
    filter(left_year > 2000 & right_year < 1990) %>%
    arrange(desc(lift)) %>% 
    select(left_title, left_year, right_title, right_year, lift) %>% 
    head(100)
```


Many Ghibli’s films and Japanese animations appear here. It looks like modern Japanese animations have enough power to draw viewers into their own world.

In contrast, only few Disney animations top the chart, which can be because they are watched by nearly everyone, which resulted in lower lift scores.

Quite surprised to see *Home on the Range* led viewers back to *The Little Mermaid*. Critical reception for the film is quite low. Maybe that reminded viewers of Disney’s renaissance era?


Next we calculate the number of common genres among two films.


```{r common_genres}
colnames(rules_df)
```


```{r common_genres_exploration}
samp <- rules_df %>% 
    sample_n(10)

# 1 is rowwise
apply(samp, 1, function(x){
    as.numeric(x[12:29])
})

apply(samp, 1, function(x) {
    as.numeric(x[34:51])
})

apply(samp, 1, function(x){
    as.numeric(x[12:29]) & as.numeric(x[34:51])
})

apply(samp, 1, function(x){
    sum(as.numeric(x[12:29]) & as.numeric(x[34:51]))
})

rm(samp)
```


```{r common_genres}
rules_df$common_genre <- apply(rules_df, 1, function(x){
    sum(as.numeric(x[12:29]) & as.numeric(x[34:51]))
})
```


```{r save_and_load_rules_df}
write_rds(rules_df, '../staging/rules.rds')
rules_df <- read_rds('../staging/rules.rds')
```


Then, we mine for a movie that led viewers to a totally different kind of movie (i.e `common_genre = 0`).

We will stick to modern films which span across different years.


```{r different_movies, paged.print=TRUE}
library(stringdist)
rules_df <- rules_df %>% 
    mutate(recent = left_year > 2000 & right_year > 2000)

rules_df %>% 
    filter(common_genre == 0, recent == TRUE, abs(left_year - right_year) >= 5) %>% 
    arrange(desc(lift)) %>% 
    select(left_title, left_year, right_title, right_year, lift)

```


Can we identify sequels

```{r heuristics}
assoc_rules = assoc_rules %>%
    filter(!(
        grepl("[0-9]", left.title, perl = TRUE) &
            grepl("[0-9]", right.title, perl = TRUE)
    )) %>%
    filter(!(
        grepl("Thin Man", left.title, perl = TRUE) &
            grepl("Thin Man", right.title, perl = TRUE)
    )) %>%
    filter(substr(
        left.title,
        start = 1,
        stop = min(5, str_length(left.title), str_length(right.title))
    ) != substr(
        right.title,
        start = 1,
        stop = min(5, str_length(left.title), str_length(right.title))
    )) %>%
    arrange(desc(lift))

head(
    assoc_rules %>% select(
        left.title,
        left.year,
        right.title,
        right.year,
        support,
        confidence,
        lift
    ),
    10
)
```


```{r}
rules_df <- rules_df %>% 
    mutate(
        movie_dist = stringdist::stringdist(left_title, right_title, method='lv')
    )

rules_df %>% 
    filter(movie_dist > 0) %>% 
    arrange(movie_dist) %>% 
    select(left_title, right_title, movie_dist) %>% 
    head(100)
```



Lastly, we can use association rules to recommend a potential movie

```{r define_movie_search}
movie_search <- function(title) {
    rules_df %>% 
    filter(str_detect(left_title, title) | str_detect(right_title, title)) %>% 
    arrange(desc(lift)) %>% 
    select(left_title, left_year, right_title, right_year, support, confidence, lift, common_genre)
}
```



```{r use_movie_search, paged.print=TRUE}
movie_search('Let the Right One In')
```



