---
title: "SVD with Users and Movies"
output:
    html_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)
library(tidyverse)
library(lubridate)
```

```{r, include=F}
closest <- function(xv, sv){
    # Returns the element in xv that is closest to the scalar value sv
    distances <- abs(xv-sv)
    xv[which(distances == min(distances))]
}

xv <- rnorm(1000, 100, 10)
# which value is closest to 108?
closest(xv, 72)
```


## Covariance


```{r load_orders}
orders <- readr::read_tsv(
    '../data/svd_orders.tsv',
    col_types = cols(
        customer=col_character(),
        date=col_date(format='%m/%d/%Y'),
        amount=col_integer()
    )
)
```

```{r orders_wide}
orders_wide <- orders %>% 
    pivot_wider(names_from='date', values_from='amount', values_fill=0)
orders_wide
```

```{r orders_matrix}
orders_matrix <- orders %>% 
    mutate(date=strftime(date, '%a')) %>% 
    rename(day=date) %>% 
    pivot_wider(names_from='day', values_from='amount', values_fill=0) %>% 
    column_to_rownames('customer') %>% 
    data.matrix()
orders_matrix
```
```{r orders_covariance_matrix}
N <- nrow(orders_matrix)
A <- t(orders_matrix) %*% orders_matrix
A
```


```{r orders_eigen_values}
spectrum <- eigen(A, symmetric=T)
round(spectrum$values, 2)
```

```{r orders_covariance_by_hand}
# covariance (the expectation of the product x*y)
# subtract column means
(col_means <- apply(orders_matrix, 2, mean))
M <- sweep(orders_matrix, 2, col_means)
(t(M) %*% M) / (N-1)
```


```{r orders_covariance}
# or more succintly
cov(orders_matrix)
friday <- orders_matrix[,3]
saturday <- orders_matrix[,4]
# remember how to compute covariance
sum((friday - mean(friday)) * (saturday - mean(saturday))) / (N - 1)
cov(friday, saturday)
# or if the matrix is already centered
friday_c <- M[,3]
saturday_c <- M[,4]
sum(friday_c * saturday_c) / (N-1)
```
```{r orders_correlation}
# if you divide by the geometric mean of the two variances you get
# the correlation coefficient
cov(friday, saturday) / sqrt(var(friday) * var(saturday))
cor(friday, saturday)
```
```{r orders_cleanup}
# clean environment
rm(list=ls())
```


## Users to Movies

$$
A = U \Sigma V^T
$$

Where $A=m \times n$, $U=m \times r$, $\Sigma=r \times r$ and $V=n \times r$


```{r load_movies}

movies <- read_tsv('../data/svd_movies.tsv', col_types = cols(
    Matrix='i',
    Alien='i',
    Serenity='i',
    Casablanca='i',
    Amelie='i'
))
```


```{r movies_matrix}
movies_matrix <- movies %>% 
    column_to_rownames('User') %>% 
    data.matrix()
movie_names <- colnames(movies_matrix)
concept_names <- c('SciFi', 'Romance', 'Noise')
movies_matrix

```

```{r movies_svd}
movies_svd <- base::svd(movies_matrix)
# how many singular vectors
round(movies_svd$d, 2)
movies_svd <- base::svd(movies_matrix, nu=3, nv=3)
U <- round(movies_svd$u, 2)
D <- diag(movies_svd$d[1:3])
V <- round(t(movies_svd$v), 2)

rownames(U) <- rownames(movies_matrix)
colnames(U) <- concept_names
colnames(V) <- movie_names
rownames(V) <- concept_names
rownames(D) <- concept_names
colnames(D) <- concept_names
```

U is _user-to-concept_ similarity matrix

```{r movies_svd_u}
U
```
Where each entry in the first column tells us how much a given user belongs to a given concept.

```{r movies_svd_s}
D
```

Each of the entries in `D` represent the strength of each concept.


```{r movies_svd_v}
V
```

V is a _movie-to-concept_ similarity matrix.

The first three movies in the first row belong to the _Sci-Fi_ concept, and the last 2 movies in the second row correspond to the _Romance_ concept.


### Reconstructing `A`

```{r movies_approx}
M <- U %*% D %*% V
colnames(M) <- colnames(movies_matrix)
rownames(M) <- rownames(movies_matrix)
round(M, 1)
```

U is row-to-concept, V is column-to-concept and $\Sigma$ models the strength of each concept.

## Dimensionality Reduction

$U \Sigma$ gives the coordinates of the points in the projection axis.

```{r movies_svd_projection}
P <- round(U %*% D, 2)
P
apply(P, 2, var)
```

## Example: Find users that like Matrix

We need to map a query into our concept space

```{r}
# remove the third singular vector
V <- V[1:2,]
```

Our vector represents a movie, so we project onto our movies-to-concepts space `V`

Remember `V` represents our movie-to-concept similarities

```{r project_movie_vectors}
concept_names <- concept_names[1:2]
# query point representing a user that likes the matrix
q <- matrix(c(5, 0, 0, 0, 0), nrow=1, dimnames=list(c('q'), movie_names))
q
# the two singular vectors in V are orthogonal to each other
# 1x5 5x2
q_concept <- q %*% t(V)
colnames(q_concept) <- concept_names
# coordinates in the concept space
abs(q_concept)
# now we have a user who likes alien and serenity
d <- matrix(c(0, 4, 5, 0, 0), nrow=1, dimnames=list(c('d'), movie_names))
d
d_concept <- d %*% t(V)
colnames(d_concept) <- concept_names
abs(d_concept)
```

Notice that the `q` and `d` had zero ratings in common, but the concept matrices are similar.


## Eigen-decomposition

SVD is

$$
A = U \Sigma V^T
$$
This relates to the diagonalization of a symmetric matrix:

$$
A = X \Lambda X^T
$$


This works when A is symmetric. `U`, `V` and `X` are orthonormal (i.e $X^T X = I$), and $\Sigma$ and $\Lambda$ are diagonal.

To create a symmetric matrix you need $A A^T$ or $A^T A$

```{r}
movies_matrix %*% t(movies_matrix)
t(movies_matrix) %*% movies_matrix
```

$$
A A^T = U \Sigma V^T (U \Sigma V^T)^T = U \Sigma V^T (V \Sigma^T U^T)
$$

Since $V^TV = I$ we get

$$
A A^T = U \Sigma \Sigma^T U^T
$$

There is a similar logic for $A^T A$

$$
A^TA = V \Sigma^T U^T (U \Sigma V^T) = V \Sigma \Sigma^T V^T
$$

And $\Sigma \Sigma^T = \Lambda$, so $\Sigma$ must contain the square roots of the eigen values, and `U` is the set of eigen vectors for $A A^T$ and `V` are the eigen values for $A^T A$


```{r}
A <- t(movies_matrix) %*% movies_matrix
res <- base::eigen(A)
round(res$values, 2)
D^2
round(res$vectors, 2)
V
```

Pros:

#. Optimal low-rank approximation in terms of the frobenius norm

Cons:

#. Interpretability problem: A singular vector specifies a linear combination of all input columns and rows.
#. Lack of Sparsity: Singular vectors are dense


```{r add_concept_group_feature_to_users}
movies$group = factor(c(1, 1, 1, 1, 2, 2, 2), labels=c('Sci-Fi', 'Romance'))
```



```{r matrix_alien_plot}
ggplot(movies, aes(x=Matrix, y=Alien, color=group)) +
    geom_point()
```

